---
title: "Getting Started with m6Aswitch"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with m6Aswitch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Getting Started with m6Aswitch

## Overview

m6Aswitch integrates isoform-level switching analysis with m6A methylation site-level quantification. This vignette walks through a typical analysis workflow.

## Installation

```r
# Install from GitHub
remotes::install_github("Tarunchikatipalli6/m6Aswitch")
```

## Load the Package

```{r setup}
library(m6Aswitch)
library(data.table)
```

## Step 1: Prepare Your Data

You need three inputs:

1. **m6A predictions** from m6AnetAnalyzer
2. **Isoform switches** from IsoformSwitchAnalyzeR
3. **Transcript sequences** for DRACH annotation

Create example data:

```{r example_data}
# Example m6A sites
m6a_example <- data.table(
  transcript_id = c("ENST001", "ENST001", "ENST002", "ENST002"),
  position = c(100, 250, 100, 150),
  probability = c(0.95, 0.72, 0.88, 0.65)
)

# Example isoform switches
switches_example <- data.table(
  gene_id = c("GENE1", "GENE2"),
  isoform_a = c("ENST001", "ENST003"),
  isoform_b = c("ENST002", "ENST004"),
  fdr = c(0.01, 0.03)
)

# Example sequences
sequences_example <- data.table(
  isoform_id = c("ENST001", "ENST002", "ENST003", "ENST004"),
  sequence = c(
    "ACUGAACUGACUGACUGACUGACUGA",
    "ACUGAACUGACUGACUGACUGACUGA",
    "GAACUGACUGACUGACUGACUGACUG",
    "GAACUGACUGACUGACUGACUGACUG"
  )
)
```

## Step 2: Parse and Integrate

```{r integrate}
# In practice, you'd load from files:
# m6a_sites <- parse_m6anet("m6anet_output.csv")
# iso_switches <- parse_isoform_switch("isoform_switches.txt")

# For this example, use the prepared data:
m6a_sites <- m6a_example
iso_switches <- switches_example

# Run the core integration
m6a_switches <- annotate_m6a_switches(m6a_sites, iso_switches, sequences_example)

head(m6a_switches)
```

## Step 3: Add DRACH Motif Annotation

```{r drach}
m6a_switches <- annotate_drach(m6a_switches, sequences_example)

# View results
m6a_switches[, .(gene_id, position, m6a_fate, drach_motif_a)]
```

## Step 4: Visualize

```{r plot}
# Summary plot of m6A fates
plot_m6a_switches(m6a_switches, plot_type = "summary")
```

## Step 5: Export Results

```{r export}
# Export to CSV and BED
results_files <- export_annotated_switches(m6a_switches, 
                                          output_prefix = "my_m6a_analysis",
                                          format = "csv")

# Generate report
report <- generate_summary_report(m6a_switches)
```

## Interpreting Results

The `m6a_fate` column describes what happens to each m6A site:

- **LOST**: Present in isoform A (normal/reference), absent in isoform B (switch). 
  - Functional impact: Loss of regulatory control
  
- **GAINED**: Absent in isoform A, present in isoform B (switch).
  - Functional impact: Acquisition of new regulation
  
- **RETAINED**: Present in both isoforms.
  - Functional impact: Consistent regulation across switch

## Next Steps

- Overlap m6A-altered genes with DESeq2 differential expression results
- Perform GO enrichment on genes with lost m6A sites
- Visualize in IGV using exported BED tracks
- Validate top hits with orthogonal methods (RIP-seq, SLAM-seq)

## Questions?

See the package documentation and GitHub repository for more details.
