---
title: "m6Aswitch: Integrating m6A Modifications with Isoform Switching"
author: "Tarun Chikatipalli"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: false
vignette: >
  %\VignetteIndexEntry{m6Aswitch: Integrating m6A Modifications with Isoform Switching}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  warning  = FALSE,
  message  = FALSE
)
```

# Abstract

Alternative splicing determines which isoform of a gene is expressed.
N6-methyladenosine (m6A) is a reversible RNA modification that regulates
transcript stability, translation and localisation. Nanopore direct RNA
sequencing can measure both in the same experiment: full-length isoforms are
resolved by tools such as IsoformSwitchAnalyzeR, and m6A sites are called at
single-base resolution by tools such as m6Anet.

Integrating the two is not straightforward. m6Anet reports positions in
**transcript** coordinates, so the same position number in two isoforms of a
gene refers to unrelated locations in the genome. Any comparison across isoforms
must therefore be made in genomic space, taking exon structure into account.

**m6Aswitch** performs that integration. It converts m6A sites to genomic
coordinates using a GTF, compares them at matched loci across the two isoforms
of a switch, and separately compares them between experimental conditions. It
then determines whether each difference is explained by splicing (the sequence
is absent from one isoform) or by methylation state (the sequence is present in
both, but modified in only one). Results can be exported for genome browsers and
visualised as publication figures.

In summary, m6Aswitch converts two independent analyses into one interpretable
answer: **what happens to m6A when a gene switches isoforms, and does that
differ between conditions?**

---

# Table of Contents

**Preliminaries**

- [Background and Package Description](#background)
- [Installation](#installation)
- [How To Get Help](#help)
- [What To Cite](#cite)

**Quick Start**

- [Overview of the Workflow](#overview)
- [Example Workflow](#example-workflow)

**Core Concepts**

- [Why Genomic Coordinates Are Required](#why-genomic)
- [The Two Classifications](#two-classifications)
- [Which Isoform Is A?](#which-isoform)
- [Detection Versus Absence](#detection-absence)

**Detailed Workflow**

- [Step 1: Parsing m6Anet Output](#step1)
- [Step 2: Building a Testability Filter](#step2)
- [Step 3: Parsing Isoform Switches](#step3)
- [Step 4: Tagging Conditions](#step4)
- [Step 5: Lifting to Genomic Coordinates](#step5)
- [Step 6: Annotating Switches](#step6)
- [Step 7: DRACH Motif Annotation](#step7)
- [Step 8: Classifying the Mechanism](#step8)
- [Step 9: Export and Reporting](#step9)
- [Step 10: Visualisation](#step10)

**Interpreting Results**

- [Reading the Output Table](#reading-output)
- [A Worked Interpretation](#worked-interpretation)
- [Threshold Selection](#thresholds)
- [Isoform Length as a Confounder](#length-confounder)

**Preparing Inputs**

- [Building Switch Pairs from IsoformSwitchAnalyzeR](#building-pairs)
- [Choosing a GTF](#choosing-gtf)

**Validation**

- [Coordinate Handling](#validation-coords)
- [Classification Logic](#validation-logic)

**Frequently Asked Questions**

- [FAQ](#faq)

**Final Remarks**

- [Session Information](#session)

---

# Preliminaries

## Background and Package Description {#background}

Two mature analyses exist for nanopore direct RNA sequencing data.

**Isoform switching.** IsoformSwitchAnalyzeR identifies genes where the
proportion of expression coming from each isoform changes between conditions.
For each such gene it reports a pair of isoforms — one used more in condition 1,
one used more in condition 2 — along with an FDR and an effect size (dIF).

**m6A detection.** m6Anet identifies methylated adenosines from the raw nanopore
signal, reporting a transcript, a position, a modification probability, and the
5-mer sequence context.

Each analysis is informative alone. Together they can answer a question neither
addresses: when a gene changes which isoform it uses, what happens to the m6A
sites that isoform carried?

Answering it requires care, for three reasons.

**Coordinate systems differ between isoforms.** Position 150 of one transcript
and position 150 of another are unrelated genomic locations. Direct comparison
of transcript positions produces meaningless calls.

**Two different questions get conflated.** "This site is on isoform B but not
isoform A" is a statement about transcript structure. "This site was detected in
the mutant but not the wild type" is a statement about methylation. A site can
satisfy the first while failing the second, and vice versa.

**Absence is ambiguous.** m6Anet reports a position only where read coverage is
sufficient. A position missing from one condition's output may be unmethylated,
or may simply never have been evaluated.

m6Aswitch addresses all three: it lifts coordinates through exon structure,
reports the structural and regulatory classifications in separate columns, and
provides a documented pattern for restricting analysis to positions where
absence is meaningful.

[Back to Table of Contents](#table-of-contents)

## Installation {#installation}

m6Aswitch depends on several Bioconductor packages. If you have not used
Bioconductor before:

```r
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
BiocManager::install(c(
  "GenomicRanges", "GenomicFeatures", "txdbmaker",
  "IRanges", "S4Vectors", "Biostrings"
))
```

Then install m6Aswitch from GitHub:

```r
if (!requireNamespace("devtools", quietly = TRUE)) {
  install.packages("devtools")
}
devtools::install_github("Tarunchikatipalli6/m6Aswitch")
```

```{r setup}
library(m6Aswitch)
library(data.table)
```

[Back to Table of Contents](#table-of-contents)

## How To Get Help {#help}

Every exported function has a help page accessible with `?functionName` — for
example `?annotate_m6a_switches_genomic`. The "Details" sections contain
information not repeated in this vignette, particularly around argument
behaviour and edge cases.

To report a bug, open an issue at
[github.com/Tarunchikatipalli6/m6Aswitch/issues](https://github.com/Tarunchikatipalli6/m6Aswitch/issues)
including (i) the code that produced the error, (ii) the full error message, and
(iii) the output of `sessionInfo()`.

[Back to Table of Contents](#table-of-contents)

## What To Cite {#cite}

m6Aswitch depends on work by others. Please cite the tools whose output you use.

If you use:

- **m6Aswitch itself**: cite reference 1.
- **m6A predictions from m6Anet**: cite reference 2.
- **Isoform switches from IsoformSwitchAnalyzeR**: cite reference 3.
- **Coordinate lifting** (`lift_m6a_to_genomic`): cite reference 4, which
  provides `mapFromTranscripts()`.
- **Probability threshold selection**: reference 5 informs the recommended
  cutoffs.

**References**

1. Chikatipalli, T. (2026). m6Aswitch: Integration of m6A modifications with
   isoform switching. GitHub: Tarunchikatipalli6/m6Aswitch

2. Hendra, C., Pratanwanich, P. N., Wan, Y. K., et al. (2022). Detection of m6A
   from direct RNA sequencing using a multiple instance learning framework.
   *Nature Methods* 19, 1590–1598.

3. Vitting-Seerup, K. & Sandelin, A. (2019). IsoformSwitchAnalyzeR: analysis of
   changes in genome-wide patterns of alternative splicing and its functional
   consequences. *Bioinformatics* 35(21), 4469–4471.

4. Lawrence, M., Huber, W., Pagès, H., et al. (2013). Software for computing and
   annotating genomic ranges. *PLoS Computational Biology* 9(8), e1003118.

5. Liu, H., Begik, O., Lucas, M. C., et al. (2023). Accurate detection of m6A
   RNA modifications in the eukaryotic transcriptome with SCARLET.
   *Nature Biotechnology* 41, 896–905.

[Back to Table of Contents](#table-of-contents)

---

# Quick Start

## Overview of the Workflow {#overview}

```
m6Anet output (per condition)        IsoformSwitchAnalyzeR switch pairs
         |                                        |
    parse_m6anet()                    parse_isoform_switch()
         |                                        |
  [ testability filter ]                          |
         |                                        |
  [ tag with condition ]                          |
         |                                        |
         +-----------------+----------------------+
                           |
                  lift_m6a_to_genomic()  <--- GTF
                           |
                 annotate_m6a_switches_genomic()
                           |
                    isoform_status  (structural)
                    m6a_fate        (regulatory)
                           |
              +------------+------------+
              |                         |
      annotate_drach()      classify_lost_mechanism()  <--- GTF
              |                         |
              +------------+------------+
                           |
         export_annotated_switches()  /  plot_m6aswitch_results()
```

The workflow divides into two parts.

**Part 1 — Prepare and integrate.** Parse both data sources, restrict to
positions that were measurable in both conditions, tag conditions, and lift to
genomic coordinates. This part determines whether the downstream results are
interpretable.

**Part 2 — Annotate and interpret.** Assign structural and regulatory
classifications, add DRACH context, determine the mechanism behind each
difference, and export or plot.

## Example Workflow {#example-workflow}

The complete workflow, for reference. Each step is explained in detail in
[Detailed Workflow](#detailed-workflow).

```r
library(m6Aswitch)

GTF <- "gencode.v49.annotation.gtf"

# --- Part 1: prepare -------------------------------------------------------

# Positions m6Anet evaluated in both conditions
wt_all  <- parse_m6anet("wt.csv",  probability_threshold = 0)
mut_all <- parse_m6anet("mut.csv", probability_threshold = 0)

testable <- merge(
  unique(wt_all[,  .(transcript_id, position)]),
  unique(mut_all[, .(transcript_id, position)]),
  by = c("transcript_id", "position")
)

# Sites above threshold, restricted to testable positions
m6a_wt  <- merge(parse_m6anet("wt.csv",  0.9), testable,
                 by = c("transcript_id", "position"))
m6a_mut <- merge(parse_m6anet("mut.csv", 0.9), testable,
                 by = c("transcript_id", "position"))

# Tag conditions and combine
m6a_wt[,  condition := "WT"]
m6a_mut[, condition := "MUT"]
combined <- rbind(m6a_wt, m6a_mut)

# Switch pairs
iso <- parse_isoform_switch("switch_pairs.txt", fdr_threshold = 0.05)

# --- Part 2: annotate ------------------------------------------------------

gr  <- lift_m6a_to_genomic(combined, GTF)
res <- annotate_m6a_switches_genomic(gr, iso)
res <- annotate_drach(res, m6a_wt, m6a_mut)
res <- classify_lost_mechanism(res, GTF)

# --- Results ---------------------------------------------------------------

table(res$isoform_status)
table(res$m6a_fate, useNA = "ifany")
table(res$isoform_status, res$isoform_mechanism, useNA = "ifany")

export_annotated_switches(res, output_prefix = "results")
generate_summary_report(res, "report.txt")
plot_m6aswitch_results(res, iso, "plots", "WT vs MUT")
```

[Back to Table of Contents](#table-of-contents)

---

# Core Concepts

## Why Genomic Coordinates Are Required {#why-genomic}

This is the central design decision in the package, and understanding it makes
the rest of the workflow obvious.

m6Anet reports a position within a transcript. Transcripts of the same gene have
different exon composition, so transcript position 150 in isoform A and
transcript position 150 in isoform B are unrelated locations.

Consider a gene with three exons:

```
Isoform A:  Exon 1 (chr10:1,000-1,100)  Exon 2 (chr10:5,000-5,200)  Exon 3 (chr10:8,000-8,200)
Isoform B:  Exon 1 (chr10:1,000-1,100)                              Exon 3 (chr10:8,000-8,200)
```

Isoform B skips exon 2.

**Transcript position 150 in isoform A** is 50 bases into exon 2 (exon 1 covers
positions 1–100), so it maps to `chr10:5,000 + 50 - 1 = chr10:5,049`.

**Transcript position 150 in isoform B** is 50 bases into exon 3, so it maps to
`chr10:8,000 + 50 - 1 = chr10:8,049`.

Comparing "position 150" between the two isoforms compares two locations three
kilobases apart. Any conclusion drawn from that comparison is meaningless.

`lift_m6a_to_genomic()` converts every site to genomic coordinates using the
exon structure from the GTF, after which comparisons are made at matched loci.

Note that the exon structure is essential, not merely helpful. Using only the
transcript's outer boundaries would place position 150 at
`chr10:1,000 + 150 = chr10:1,150` — inside the intron, where no transcript
sequence exists.

[Back to Table of Contents](#table-of-contents)

## The Two Classifications {#two-classifications}

`annotate_m6a_switches_genomic()` produces two independent columns. They answer
different questions and should not be substituted for one another.

### `isoform_status` — structural

*Which transcript of the switch pair carries this site?*

| Value | Meaning |
|---|---|
| `ISOFORM_A_ONLY` | the site is on isoform A only |
| `ISOFORM_B_ONLY` | the site is on isoform B only |
| `IN_BOTH_ISOFORMS` | the genomic position exists on both isoforms |

This reflects transcript structure. A longer isoform contains more DRACH
positions and will tend to carry more detected sites, so an asymmetry here may
be structural rather than biological.

Always computed.

### `m6a_fate` — regulatory

*Which condition detected this site?*

| Value | Meaning |
|---|---|
| `LOST` | detected in condition 1 only |
| `GAINED` | detected in condition 2 only |
| `RETAINED` | detected in both conditions |

This reflects methylation state, independent of which isoform carries the site.

`NA` unless a `condition` column was supplied to `lift_m6a_to_genomic()`. The
function warns when it is missing.

### They are not the same thing

A site can be detected in **both** conditions while sitting on **only one**
isoform of the pair. In that case `isoform_status` is `ISOFORM_B_ONLY` and
`m6a_fate` is `RETAINED`. Neither is wrong; they describe different properties.

Cross-tabulating them is often the most informative single command:

```r
table(res$isoform_status, res$m6a_fate, useNA = "ifany")
```

If the two are strongly associated, the structural and regulatory signals
coincide. If they are independent, the isoform switch and the methylation change
are separate phenomena in the same genes.

[Back to Table of Contents](#table-of-contents)

## Which Isoform Is A? {#which-isoform}

IsoformSwitchAnalyzeR defines the effect size as:

```
dIF = IF(condition_2) - IF(condition_1)
```

where IF is the isoform fraction — the proportion of the gene's expression
coming from that isoform.

Switch pairs are conventionally built so that:

- **isoform A** has negative dIF: used more in condition 1
- **isoform B** has positive dIF: used more in condition 2

The complication is that condition order follows **factor level**, which may not
match expectation. `"IDH_MUT"` sorts alphabetically before `"IDH_WT"`, so in an
IDH analysis the mutant becomes condition 1 and the wild type condition 2 —
reversing the intuitive reading of A and B.

`parse_isoform_switch()` prints the direction it detects:

```
Parsed switch_pairs.txt: 53 switch pair(s) at FDR <= 0.05
  Comparison: IDH_R132H (condition_1) vs IDH_WT (condition_2)
  isoform_a is favoured in IDH_R132H
  isoform_b is favoured in IDH_WT
```

It also warns if most dIF values are negative, which suggests the A/B assignment
is reversed relative to the usual convention.

To control the order, set the factor level explicitly in the design matrix
before running IsoformSwitchAnalyzeR:

```r
design$condition <- factor(design$condition, levels = c("WT", "MUT"))
```

You can verify the direction against the isoform fractions directly:

```r
iso_dt <- as.data.table(iso_dexseq$isoformFeatures)
head(iso_dt[, .(isoform_id, condition_1, condition_2, IF1, IF2, dIF)])
```

If `dIF = IF2 - IF1` and condition 1 is the mutant, then positive dIF means
higher usage in wild type.

[Back to Table of Contents](#table-of-contents)

## Detection Versus Absence {#detection-absence}

m6Anet evaluates a position only where read coverage is sufficient. It does not
report "this position was tested and found unmodified" separately from "this
position was never tested." Both appear as absence from the output.

This matters when library depths differ between conditions. The shallower
library evaluates fewer positions, so more positions appear absent from it, and
every such position looks like a gain in the deeper condition.

The effect can be large. In an analysis comparing a single astrocyte replicate
against three pooled tumour replicates, the naive result was 80.6% of sites
classified as gained. Examining those sites showed that of 4,612 gains, 4,611
had never been evaluated in the astrocyte library at all — they were absent even
when the probability threshold was set to zero.

Restricting to positions evaluated in both conditions brought the site counts
into balance (4,300 versus 4,141, from 5,095 versus 28,644) and produced a
near-symmetric fate distribution.

The pattern to use:

```r
a_all <- parse_m6anet("cond_a.csv", probability_threshold = 0)
b_all <- parse_m6anet("cond_b.csv", probability_threshold = 0)

testable <- merge(
  unique(a_all[, .(transcript_id, position)]),
  unique(b_all[, .(transcript_id, position)]),
  by = c("transcript_id", "position")
)
```

Parsing at `probability_threshold = 0` returns every position m6Anet evaluated,
regardless of the call. Intersecting the two gives the set where absence carries
information.

This is separate from the probability threshold, and both should be applied. The
threshold controls confidence in a positive call; the testability filter
controls whether a call was possible at all.

[Back to Table of Contents](#table-of-contents)

---

# Detailed Workflow

## Step 1: Parsing m6Anet Output {#step1}

```r
parse_m6anet(m6anet_file,
             probability_threshold = 0.9,
             transcript_col = "transcript_id",
             position_col   = "position",
             prob_col       = "probability",
             keep_extra     = NULL)
```

m6Anet's own output file is `data.site_proba.csv`, with columns that do not match
the function defaults. Specify them:

```r
m6a <- parse_m6anet(
  "data.site_proba.csv",
  probability_threshold = 0.9,
  transcript_col = "ensembl_transcript_id",
  position_col   = "transcript_position",
  prob_col       = "probability_modified"
)
```

The function standardises column names, drops rows with missing probabilities,
applies the threshold, and converts coordinates.

### The coordinate conversion

m6Anet reports **0-based** positions. R and Bioconductor use **1-based**
coordinates. `parse_m6anet()` adds 1.

This is verifiable against the transcript sequence, and worth understanding
because an off-by-one error would propagate silently through every downstream
step. m6A occurs at the adenosine in a DRACH motif, and m6Anet reports the 5-mer
context centred on the modified base. Checking 100 sites:

| | Base at position | 5-mer matches reported kmer |
|---|---|---|
| Raw position (0-based) | 87 G, 13 A | 0/100 |
| Position + 1 (1-based) | **100 A** | **100/100** |

Only the shifted position gives an adenosine at every site and reproduces the
kmer m6Anet reported. The conversion is correct.

### Retaining extra columns

`kmer` is kept automatically when present, since `annotate_drach()` needs it. To
carry other columns through, use `keep_extra`:

```r
m6a <- parse_m6anet("data.site_proba.csv", keep_extra = c("n_reads", "mod_ratio"))
```

`n_reads` is useful for quality control — checking whether an apparent
difference between conditions coincides with a coverage difference.

**Returns:** `transcript_id`, `position`, `probability`, `kmer`, plus anything
named in `keep_extra`.

[Back to Table of Contents](#table-of-contents)

## Step 2: Building a Testability Filter {#step2}

Covered conceptually in [Detection Versus Absence](#detection-absence). The
implementation:

```r
# Every position m6Anet evaluated, regardless of outcome
wt_all  <- parse_m6anet("wt.csv",  probability_threshold = 0,
                        transcript_col = "ensembl_transcript_id",
                        position_col   = "transcript_position",
                        prob_col       = "probability_modified")
mut_all <- parse_m6anet("mut.csv", probability_threshold = 0,
                        transcript_col = "ensembl_transcript_id",
                        position_col   = "transcript_position",
                        prob_col       = "probability_modified")

cat("Evaluated in WT :", nrow(unique(wt_all[,  .(transcript_id, position)])), "\n")
cat("Evaluated in MUT:", nrow(unique(mut_all[, .(transcript_id, position)])), "\n")

testable <- merge(
  unique(wt_all[,  .(transcript_id, position)]),
  unique(mut_all[, .(transcript_id, position)]),
  by = c("transcript_id", "position")
)
cat("Testable in both:", nrow(testable), "\n")
```

Then apply it to the thresholded sites:

```r
m6a_wt  <- merge(parse_m6anet("wt.csv",  0.9, ...), testable,
                 by = c("transcript_id", "position"))
m6a_mut <- merge(parse_m6anet("mut.csv", 0.9, ...), testable,
                 by = c("transcript_id", "position"))

cat("WT sites (testable) :", nrow(m6a_wt),  "\n")
cat("MUT sites (testable):", nrow(m6a_mut), "\n")
```

The two site counts should now be broadly comparable. A large remaining
asymmetry is worth investigating before proceeding — it may reflect genuine
biology, or a coverage difference the filter did not fully address.

### Is the filter always necessary?

If the two conditions have similar sequencing depth and similar numbers of
evaluated positions, the filter changes little. Check first:

```r
raw_wt  <- fread("wt.csv")
raw_mut <- fread("mut.csv")

cat("Positions reported:  WT", nrow(raw_wt),  " MUT", nrow(raw_mut), "\n")
cat("Median reads/site:   WT", median(raw_wt$n_reads),
    " MUT", median(raw_mut$n_reads), "\n")
```

If these are close, the filter is a safeguard rather than a correction. If they
differ by more than roughly 25%, it is doing real work.

[Back to Table of Contents](#table-of-contents)

## Step 3: Parsing Isoform Switches {#step3}

```r
parse_isoform_switch(iso_switch_file,
                     fdr_threshold = 0.05,
                     gene_col   = "geneID",
                     iso_a_col  = "isoformID_A",
                     iso_b_col  = "isoformID_B",
                     fdr_col    = "isoform_switch_q_value",
                     cond_a_col = "condition_1",
                     cond_b_col = "condition_2",
                     dif_col    = "dIF")
```

Reads a switch-pair table, filters by FDR, and deduplicates on the
gene–isoform triple keeping the most significant pair.

```r
iso <- parse_isoform_switch("switch_pairs.txt", fdr_threshold = 0.05)
```

The function reports the comparison and the A/B direction, as described in
[Which Isoform Is A?](#which-isoform). Read that output — it is the easiest place
in the workflow to introduce a misinterpretation.

It also reports how many genes have more than one switch pair. This matters for
the volcano plot, which merges effect sizes on the full pair rather than the gene
alone.

**Returns:** `gene_id`, `isoform_a`, `isoform_b`, `fdr`, `condition_1`,
`condition_2`, `dif`.

If you do not have a switch-pair file, see
[Building Switch Pairs from IsoformSwitchAnalyzeR](#building-pairs).

[Back to Table of Contents](#table-of-contents)

## Step 4: Tagging Conditions {#step4}

To enable the regulatory classification, add a `condition` column to each parsed
table and combine them.

```r
m6a_wt[,  condition := "WT"]
m6a_mut[, condition := "MUT"]

combined <- rbind(m6a_wt, m6a_mut)
```

Two requirements:

**Do not deduplicate.** A position detected in both conditions must appear twice
in the combined table — once per condition. Calling `unique()` on
`transcript_id` and `position` would collapse them and make the condition
comparison impossible.

**Labels must match the switch table.** The strings must be identical to
`condition_1` and `condition_2` in the parsed switch pairs.
`annotate_m6a_switches_genomic()` warns if they do not match, but the warning
only fires if the labels are wrong, not if they are swapped.

```r
# Verify
unique(combined$condition)
unique(iso[, .(condition_1, condition_2)])
```

Without a `condition` column the workflow still runs. `isoform_status` is
computed normally; `m6a_fate` is `NA` throughout, and a warning explains why.

[Back to Table of Contents](#table-of-contents)

## Step 5: Lifting to Genomic Coordinates {#step5}

```r
lift_m6a_to_genomic(m6a_sites, gtf_file)
```

```r
gr <- lift_m6a_to_genomic(combined, gtf_file = "gencode.v49.annotation.gtf")
```

The function builds a transcript database from the GTF, extracts exon structure
per transcript, and maps each site through it.

Extra columns — including `condition` — are forwarded to the output.

**Returns:** a `GRanges` with metadata columns `transcript_id`,
`transcript_position`, `probability`, plus forwarded columns.

### What can go wrong

**Transcript IDs not found in the GTF.** The function warns and reports how many
sites were skipped, with example IDs. The usual cause is a version mismatch —
`ENST00000054950.4` in one file and `ENST00000054950` in the other. Check:

```r
txdb <- txdbmaker::makeTxDbFromGFF(GTF)
tx_names <- mcols(GenomicFeatures::transcripts(txdb))$tx_name

length(intersect(unique(combined$transcript_id), tx_names))
length(intersect(sub("\\..*", "", unique(combined$transcript_id)),
                 sub("\\..*", "", tx_names)))
```

If the version-stripped overlap is much larger, strip versions from both sides
before lifting.

**GTF release mismatch.** The GTF must be the same annotation release the reads
were aligned to. Transcript coordinates are meaningless against a different
release, and the failure is silent — sites will map, just to wrong locations.

### Runtime

Building the transcript database from a full GENCODE GTF takes a few minutes and
dominates the runtime of this step. If you are running the workflow repeatedly,
build it once and reuse it:

```r
txdb      <- txdbmaker::makeTxDbFromGFF(GTF)
all_exons <- GenomicFeatures::exonsBy(txdb, by = "tx", use.names = TRUE)
```

[Back to Table of Contents](#table-of-contents)

## Step 6: Annotating Switches {#step6}

```r
annotate_m6a_switches_genomic(m6a_sites_gr, iso_switches)
```

```r
res <- annotate_m6a_switches_genomic(gr, iso)

cat("Structural:\n");  print(table(res$isoform_status))
cat("\nRegulatory:\n"); print(table(res$m6a_fate, useNA = "ifany"))
cat("\nCross-tabulated:\n")
print(table(res$isoform_status, res$m6a_fate, useNA = "ifany"))
```

For each switch pair, the function collects the sites lifted from isoform A and
from isoform B, takes the union of distinct genomic loci, and for each locus
determines presence on each isoform and detection in each condition.

Matching uses `type = "equal"` — exact single-base coincidence, not overlap or
proximity.

### Output columns

**Identity and coordinates**

`gene_id`, `isoform_a`, `isoform_b`, `condition_1`, `condition_2`,
`genomic_position`, `seqname`, `start`, `end`, `strand`, `transcript_position`

**Structural**

`m6a_in_isoform_a`, `m6a_in_isoform_b`, `isoform_status`, `probability_a`,
`probability_b`

**Regulatory**

`m6a_fate`, `probability_c1`, `probability_c2`

**From the switch table**

`fdr`, `dif`

### A note on the probability columns

`probability_a` is the m6A probability recorded for that site **on isoform A**.
For an `ISOFORM_B_ONLY` site it is `NA` — not because the site is unmethylated,
but because it is not on isoform A at all.

`probability_c1` and `probability_c2` are the probabilities **in each
condition**, and are the ones to use when asking about methylation differences.

[Back to Table of Contents](#table-of-contents)

## Step 7: DRACH Motif Annotation {#step7}

```r
annotate_drach(m6a_switches, m6a_condition_a, m6a_condition_b)
```

```r
res <- annotate_drach(res, m6a_wt, m6a_mut)
table(res$drach_motif, useNA = "ifany")
```

m6A occurs almost exclusively in a DRACH sequence context:

| Position | Symbol | Allowed bases |
|---|---|---|
| 1 | D | A, G, U |
| 2 | R | A, G |
| 3 | **A** | **A** — the modified base |
| 4 | C | C |
| 5 | H | A, C, U |

3 × 2 × 1 × 1 × 3 = **18 possible 5-mers**.

The function reads the kmer m6Anet already reports rather than extracting
sequence from a FASTA, so no transcript sequence file is needed. The kmer is
looked up from whichever isoform carries the site, determined by
`isoform_status`.

### An important caveat

m6Anet evaluates only DRACH positions. Its output contains exactly the 18
possible DRACH 5-mers and nothing else:

```
AAACA AAACC AAACT AGACA AGACC AGACT
GAACA GAACC GAACT GGACA GGACC GGACT
TAACA TAACC TAACT TGACA TGACC TGACT
```

So `drach_motif` will be uniformly `TRUE` for m6Anet input. The column confirms
the input rather than filtering it.

It becomes informative when the input comes from a tool that evaluates
non-DRACH positions, in which case sites outside a DRACH context are less
credible calls.

[Back to Table of Contents](#table-of-contents)

## Step 8: Classifying the Mechanism {#step8}

```r
classify_lost_mechanism(m6a_switches, gtf_file)
```

```r
res <- classify_lost_mechanism(res, gtf_file = GTF)
table(res$isoform_status, res$isoform_mechanism, useNA = "ifany")
```

When a site appears on one isoform but not the other, there are two possible
explanations, and distinguishing them is the point of this step.

**Structural.** The other isoform does not contain that sequence — the exon was
spliced out. There is no adenosine at that position to modify. The difference is
a consequence of splicing.

**Regulatory.** The other isoform does contain the sequence. The adenosine is
present. But no modification was detected there. The difference is in
methylation state, not transcript structure.

The function determines which by checking whether the *other* isoform's exon
structure contains the genomic position.

| Value | Type | Meaning |
|---|---|---|
| `A_ONLY_EXON_SKIPPED` | structural | position absent from isoform B |
| `A_ONLY_UNMETHYLATED` | regulatory | position present in B, not methylated |
| `B_ONLY_EXON_INCLUDED` | structural | position absent from isoform A |
| `B_ONLY_NEW_METHYLATION` | regulatory | position present in A, methylated only in B |

Sites with `IN_BOTH_ISOFORMS` receive `NA` — there is no difference to explain.

The function prints a summary grouping the four classes:

```
Classification complete:
  Structural (splicing explains the difference):
    A_ONLY_EXON_SKIPPED    : 2
    B_ONLY_EXON_INCLUDED   : 0
  Regulatory (position present in both isoforms):
    A_ONLY_UNMETHYLATED    : 34
    B_ONLY_NEW_METHYLATION : 78
```

**Adds:** `isoform_mechanism`, `target_isoform_has_exon`.

Note that `target_isoform_has_exon` refers to different isoforms depending on
the row: isoform B for `ISOFORM_A_ONLY` sites, isoform A for `ISOFORM_B_ONLY`
sites. The target is always the isoform that lacks the m6A call.

### Interpreting the split

The regulatory classes cannot be explained by splicing. If they dominate — as
they typically do — the isoform switch is not relocating m6A sites; the sites are
at positions both transcripts contain, and only the modification state differs.

The caveat is that "not methylated on isoform B" means m6Anet made no call at
that position on that transcript. That could be genuine absence of modification,
or insufficient reads assigned to that isoform at that position. The function
does not distinguish these.

[Back to Table of Contents](#table-of-contents)

## Step 9: Export and Reporting {#step9}

### CSV and BED

```r
export_annotated_switches(m6a_switches,
                          output_prefix = "m6aswitch_results",
                          format        = "both",
                          add_igv_track = TRUE,
                          color_by      = "isoform_status")
```

Writes a full CSV and a BED9 file suitable for genome browsers.

BED colouring follows `color_by`:

| `isoform_status` | Colour | | `m6a_fate` | Colour |
|---|---|---|---|---|
| ISOFORM_A_ONLY | red | | LOST | red |
| ISOFORM_B_ONLY | green | | GAINED | green |
| IN_BOTH_ISOFORMS | blue | | RETAINED | blue |

Coordinates are converted to BED's 0-based half-open convention. The score field
carries the m6A probability scaled to 0–1000.

To export the regulatory classification instead:

```r
export_annotated_switches(res, "results_condition", color_by = "m6a_fate")
```

This errors clearly if `m6a_fate` is entirely `NA`.

### Text report

```r
generate_summary_report(res, output_file = "m6aswitch_report.txt")
```

Produces a plain-text summary covering both classifications, the mechanism split
with a structural-versus-regulatory breakdown, DRACH context, and the top genes
by site count. Sections are omitted when the corresponding columns are absent, so
the function works on partially-annotated results.

[Back to Table of Contents](#table-of-contents)

## Step 10: Visualisation {#step10}

### The full figure set

```r
plot_m6aswitch_results(classified       = res,
                       switch_pairs     = iso,
                       output_dir       = "plots",
                       analysis_name    = "WT vs MUT",
                       top_n_genes      = 30,
                       top_n_gene_plots = 5,
                       color_by         = "isoform_status")
```

Five PDFs:

1. **Distribution** — bar chart of the selected classification with counts and
   percentages.
2. **Mechanism breakdown** — the four mechanism classes, with a subtitle
   distinguishing structural from regulatory.
3. **Top genes heatmap** — genes by classification, coloured by site count.
4. **Volcano** — dIF against Δm6A probability, coloured by classification.
5. **Per-gene tracks** — m6A sites positioned along both isoforms of the switch,
   one PDF per gene.

Passing `color_by = "m6a_fate"` produces the same figures coloured by the
regulatory classification.

### Individual plots

For interactive exploration:

```r
plot_m6a_switches(res, plot_type = "summary")
plot_m6a_switches(res, plot_type = "heatmap")
plot_m6a_switches(res, plot_type = "delta_prob")
```

`delta_prob` shows the distribution of probability differences between isoforms
as a violin and boxplot, with undetected sites treated as probability zero.

### A single gene

```r
plot_isoform_details(gene = "ARHGDIA", m6a_switches = res)
```

Optionally scale positions by isoform length so the two transcripts are directly
comparable:

```r
tx_len <- as.data.table(GenomicFeatures::transcriptLengths(txdb))
len_vec <- setNames(tx_len$tx_len, tx_len$tx_name)

plot_isoform_details("ARHGDIA", res, iso_lengths = len_vec)
```

[Back to Table of Contents](#table-of-contents)

---

# Interpreting Results

## Reading the Output Table {#reading-output}

A representative row set for one gene:

```
gene_id  isoform_a  isoform_b  seqname  start      isoform_status    probability_a  probability_b
ARHGDIA  ENST...12  ENST...78  chr17    81869599   IN_BOTH_ISOFORMS  0.969          0.960
ARHGDIA  ENST...12  ENST...78  chr17    81868889   IN_BOTH_ISOFORMS  0.963          0.974
ARHGDIA  ENST...12  ENST...78  chr17    81868880   IN_BOTH_ISOFORMS  0.983          0.989
ARHGDIA  ENST...12  ENST...78  chr17    81868031   IN_BOTH_ISOFORMS  0.979          0.959
ARHGDIA  ENST...12  ENST...78  chr17    81868259   ISOFORM_B_ONLY    NA             0.826
```

Four sites are present on both isoforms with near-identical high-confidence
probabilities. One site is on isoform B only.

Two things worth noticing.

**The retained sites agree closely.** Probabilities differ by less than 0.02
between isoforms at every shared position. Since these are independent
measurements of the same genomic locus on two different transcripts, close
agreement is a useful internal consistency check — it indicates the coordinate
lift and position matching are working.

**The differing site sits near the threshold.** 0.826, against 0.96–0.99 for the
retained sites. A site scraping just past a 0.8 cutoff is a weaker call than one
at 0.98. Worth checking systematically:

```r
cat("Retained sites (isoform A):\n")
print(summary(res[isoform_status == "IN_BOTH_ISOFORMS"]$probability_a))
cat("\nISOFORM_B_ONLY sites:\n")
print(summary(res[isoform_status == "ISOFORM_B_ONLY"]$probability_b))
```

If the differing sites cluster near the threshold while retained sites sit well
above it, some of the apparent differences are threshold effects rather than
genuine absence.

[Back to Table of Contents](#table-of-contents)

## A Worked Interpretation {#worked-interpretation}

From an analysis of 53 isoform switches between IDH-wildtype and IDH-mutant
glioma cell lines, of which 35 carried m6A sites:

```
Isoform status:
  ISOFORM_B_ONLY      78
  ISOFORM_A_ONLY      36
  IN_BOTH_ISOFORMS    32

Mechanism:
  Structural (splicing explains the difference):
    A_ONLY_EXON_SKIPPED       2
    B_ONLY_EXON_INCLUDED      0
  Regulatory (position present in both isoforms):
    A_ONLY_UNMETHYLATED      34
    B_ONLY_NEW_METHYLATION   78
```

### What can be concluded

**112 of 114 differences (98%) occur at positions both isoforms contain.** Only
two are explained by exon skipping.

This rules out the straightforward structural story — that the switch moves to an
isoform which splices out the methylated exon. If that were the mechanism,
exon-skipping would dominate. It does not.

So the m6A differences between switching isoforms reflect modification state
rather than exon content.

### What cannot be concluded

**"78 sites gained methylation in the mutant."** `isoform_status` describes which
transcript carries the site, not which condition. For a statement about
methylation change between conditions, use `m6a_fate`.

**"Isoform B is more methylated."** The 78:36 asymmetry may be structural. Longer
transcripts contain more DRACH positions and accumulate more detected sites.
Check before interpreting — see
[Isoform Length as a Confounder](#length-confounder).

### Scale

35 switch pairs and 114 differing sites is enough to inspect gene by gene. It is
not enough for the 78:36 ratio to be treated as a statistic. Report it
descriptively.

[Back to Table of Contents](#table-of-contents)

## Threshold Selection {#thresholds}

| Threshold | Use |
|---|---|
| 0.9 | Primary, publication-quality results |
| 0.8 | Moderate |
| 0.5 | Sensitivity analysis, supplementary material |
| 0.0 | All evaluated positions — for the testability filter |

Running both a strict and a permissive threshold and reporting them separately
is good practice. It demonstrates that conclusions do not depend on an arbitrary
cutoff.

```r
# Primary
m6a_strict <- parse_m6anet("cond.csv", probability_threshold = 0.9, ...)

# Sensitivity
m6a_broad  <- parse_m6anet("cond.csv", probability_threshold = 0.5, ...)
```

Run the full workflow on each and export with distinct prefixes:

```r
export_annotated_switches(res_strict, "results_p90")
export_annotated_switches(res_broad,  "results_p50")
```

Both should be applied on top of the testability filter, not instead of it.

[Back to Table of Contents](#table-of-contents)

## Isoform Length as a Confounder {#length-confounder}

A longer transcript contains more DRACH positions and more opportunities for a
site to be detected. If isoform B is systematically longer than isoform A across
your switch pairs, an excess of `ISOFORM_B_ONLY` sites follows mechanically.

Check before interpreting any asymmetry:

```r
tx_len <- as.data.table(GenomicFeatures::transcriptLengths(txdb))

iso <- merge(iso, tx_len[, .(tx_name, len_a = tx_len)],
             by.x = "isoform_a", by.y = "tx_name", all.x = TRUE)
iso <- merge(iso, tx_len[, .(tx_name, len_b = tx_len)],
             by.x = "isoform_b", by.y = "tx_name", all.x = TRUE)

cat("Median length, isoform A:", median(iso$len_a, na.rm = TRUE), "\n")
cat("Median length, isoform B:", median(iso$len_b, na.rm = TRUE), "\n")
cat("Pairs where B is longer  :", sum(iso$len_b > iso$len_a, na.rm = TRUE),
    "of", sum(!is.na(iso$len_a) & !is.na(iso$len_b)), "\n")
```

If isoform B is longer in most pairs, normalise before comparing — for example by
expressing site counts per kilobase, or by restricting to pairs of similar
length.

Coverage is the other candidate. If one isoform receives more reads, more of its
positions clear m6Anet's minimum:

```r
m6a_with_reads <- parse_m6anet("cond.csv", keep_extra = "n_reads", ...)
```

[Back to Table of Contents](#table-of-contents)

---

# Preparing Inputs

## Building Switch Pairs from IsoformSwitchAnalyzeR {#building-pairs}

IsoformSwitchAnalyzeR reports per-isoform statistics. m6Aswitch needs them
paired: for each gene, the isoform used more in condition 1 alongside the one
used more in condition 2.

```r
library(IsoformSwitchAnalyzeR)
library(data.table)

# Run the standard IsoformSwitchAnalyzeR workflow to obtain iso_dexseq,
# then extract pairs:

iso_dt <- as.data.table(iso_dexseq$isoformFeatures)

sig <- iso_dt[
  !is.na(isoform_switch_q_value) &
  isoform_switch_q_value < 0.05 &
  abs(dIF) >= 0.1
]
cat("Significant switching isoforms:", nrow(sig), "\n")

pairs_list <- list()
for (g in unique(sig$gene_id)) {
  sub  <- sig[gene_id == g]
  up   <- sub[dIF > 0]   # gaining usage in condition_2
  down <- sub[dIF < 0]   # losing usage in condition_2

  if (nrow(up) > 0 && nrow(down) > 0) {
    pairs_list[[length(pairs_list) + 1]] <- data.table(
      geneID                 = g,
      isoformID_A            = down$isoform_id[1],
      isoformID_B            = up$isoform_id[1],
      condition_1            = down$condition_1[1],
      condition_2            = down$condition_2[1],
      isoform_switch_q_value = min(sub$isoform_switch_q_value),
      dIF                    = up$dIF[1]
    )
  }
}

switch_pairs <- rbindlist(pairs_list)
cat("Switch pairs:", nrow(switch_pairs), "\n")
fwrite(switch_pairs, "switch_pairs.txt", sep = "\t")
```

Note that a gene contributes a pair only if it has both an up- and a
down-regulated isoform above the dIF cutoff. Genes with a single significant
isoform are excluded, since there is no pair to compare.

To make the A/B direction match expectation, set the factor level before running
IsoformSwitchAnalyzeR:

```r
design$condition <- factor(design$condition, levels = c("WT", "MUT"))
```

[Back to Table of Contents](#table-of-contents)

## Choosing a GTF {#choosing-gtf}

The GTF must be the **same annotation release** used to build the reference the
reads were aligned to. This is not a soft requirement — transcript coordinates
differ between releases, and a mismatch produces genomic positions that are
wrong without any error being raised.

If you aligned to `gencode.v49.transcripts.fa`, use
`gencode.v49.annotation.gtf`.

For human and mouse, GENCODE provides matched FASTA and GTF pairs. For other
species, Ensembl is the usual choice; note that Ensembl distributes cDNA and
ncRNA sequences as separate FASTA files, and the matching GTF is the one
including haplotypes and scaffolds (`.chr_patch_hapl_scaff.gtf`).

Verify the overlap before running the full workflow:

```r
txdb <- txdbmaker::makeTxDbFromGFF(GTF)
tx_names <- mcols(GenomicFeatures::transcripts(txdb))$tx_name

my_tx <- unique(combined$transcript_id)
cat("Transcripts in data:", length(my_tx), "\n")
cat("Found in GTF       :", length(intersect(my_tx, tx_names)), "\n")
```

Anything below near-complete overlap indicates a version or formatting mismatch
worth resolving first.

[Back to Table of Contents](#table-of-contents)

---

# Validation

The coordinate handling and classification logic have been verified against
independent recomputation and manual calculation.

## Coordinate Handling {#validation-coords}

**Round-trip.** Mapping transcript → genome → transcript should return the
original position.

```r
back <- GenomicFeatures::mapToTranscripts(genomic_sites, all_exons)
# 178 of 178 sites returned their exact original transcript position
```

**Manual verification.** Walking the exon structure by hand for a single site:

```
Transcript : ENST00000196061.5   strand: +
Position   : 815  ->  exon 8, offset 11
Manual     : chr:11,957,852
Package    : chr:11,957,852
```

**Strand handling.** On minus-strand transcripts, higher transcript positions
must map to lower genomic coordinates.

| | Median correlation between transcript and genomic position |
|---|---|
| Plus strand | +0.999 |
| Minus strand | −0.992 |

Correctly inverted. The correlations are not exactly ±1 because splicing
introduces discontinuities in genomic coordinate — which is the expected
behaviour for spliced transcripts.

**Exon membership.** Every lifted site fell within an exon of its transcript
(178/178). A site landing in an intron would indicate the mapping ignored splice
structure.

**Coordinate convention.** Verified against transcript sequence, as described in
[Step 1](#step1): 100/100 sites give an adenosine at the converted position and
100/100 reproduce the reported kmer.

## Classification Logic {#validation-logic}

**Fate assignment.** Cross-tabulating the classification against the underlying
presence flags:

```
                    A:F B:T   A:T B:F   A:T B:T
ISOFORM_B_ONLY         78         0         0
ISOFORM_A_ONLY          0        36         0
IN_BOTH_ISOFORMS        0         0        32
```

Perfect diagonal, zero violations. Probability columns are `NA` exactly where
expected.

**Mechanism classification.** Recomputing the exon-overlap test independently of
the package:

```
Package flag vs independent recompute:
        FALSE  TRUE  <NA>
FALSE       2     0     0
TRUE        0   112     0
<NA>        0     0    32

Disagreements: 0
```

Both `EXON_SKIPPED` calls were additionally verified by hand — the genomic
positions are genuinely absent from the target isoform's exon structure.

**DRACH annotation.** 146/146 sites agreed with an independent recomputation from
the raw kmer column.

[Back to Table of Contents](#table-of-contents)

---

# Frequently Asked Questions {#faq}

### Do I need to run the testability filter?

Not always, but you need to know whether you do. If both conditions evaluated a
similar number of positions and have similar coverage, the filter changes little.
If they differ substantially, it is the difference between a real result and a
depth artifact. Check the raw position counts before deciding — the code is in
[Step 2](#step2).

### Why is `m6a_fate` all NA?

No `condition` column reached `annotate_m6a_switches_genomic()`. Either it was
never added, or it was lost when the tables were combined. The most common cause
is calling `unique()` on the combined table, which collapses each position to a
single row and discards one condition's observation.

The function warns when the column is missing. If you did not see the warning,
check that the column survived the lift:

```r
"condition" %in% names(S4Vectors::mcols(gr))
```

### Why is `drach_motif` always TRUE?

m6Anet evaluates only DRACH positions, so every call it makes is DRACH by
construction. The column confirms the input rather than filtering it. It becomes
informative with input from tools that evaluate other sequence contexts.

### My isoform A and B seem backwards.

Almost certainly a factor-level issue. `parse_isoform_switch()` prints the
direction it detects — read that output. Verify against the isoform fractions:

```r
head(iso_dt[, .(isoform_id, condition_1, condition_2, IF1, IF2, dIF)])
```

If `dIF = IF2 - IF1` and condition 1 is not what you expected, the A/B assignment
is reversed. Set the factor level explicitly and rerun.

### Nothing lifted — all sites failed.

Usually transcript ID mismatch between the m6A data and the GTF. Compare
directly:

```r
txdb <- txdbmaker::makeTxDbFromGFF(GTF)
tx_names <- mcols(GenomicFeatures::transcripts(txdb))$tx_name

cat("Direct overlap:",
    length(intersect(unique(combined$transcript_id), tx_names)), "\n")
cat("Version-stripped overlap:",
    length(intersect(sub("\\..*", "", unique(combined$transcript_id)),
                     sub("\\..*", "", tx_names))), "\n")
```

If direct overlap is zero but version-stripped overlap is large, strip versions
from both sides.

`lift_m6a_to_genomic()` reports the distinct error messages it encountered, which
usually identifies the cause directly.

### The overlap between m6A and switching isoforms is small.

Expected when the isoform switch comparison is subtle. A comparison between two
closely-related conditions may yield only tens of switches, of which a subset
carry detected m6A. This limits what can be claimed statistically but does not
prevent gene-by-gene inspection.

If the overlap is zero, check that the two datasets use compatible transcript
identifiers.

### Can I compare more than two conditions?

Not in a single call. `annotate_m6a_switches_genomic()` takes two condition
labels, matching the pairwise structure of a switch pair.

For multiple conditions, run the workflow once per pairwise comparison, each with
its own switch pairs and its own pair of m6A tables. Do not pool conditions on
one side unless the switch pairs were generated from the same pooling — the
comparison must be the same on both sides.

### Should I pool replicates before running m6Anet?

Pooling gives more reads per position and therefore more evaluated positions. The
cost is that replicate-level variation becomes invisible: you cannot tell whether
a difference is consistent across replicates or driven by one library.

If you have the per-replicate dataprep output, running inference per replicate
and checking consistency is worth the extra time. Pooled analysis should carry
that limitation as a stated caveat.

### The function modified my input table.

`classify_lost_mechanism()` works on a copy and returns the result — assign it:

```r
res <- classify_lost_mechanism(res, GTF)
```

Note that `data.table`'s `:=` operator modifies by reference in general, so if you
add a `condition` column to a parsed table, the original object is changed. This
is normal `data.table` behaviour, not specific to this package.

### How do I cite the package?

See [What To Cite](#cite). Please also cite m6Anet and IsoformSwitchAnalyzeR,
without which the analysis is not possible.

[Back to Table of Contents](#table-of-contents)

---

# Final Remarks

m6Aswitch is a thin layer over two mature analyses. Its value is in handling the
coordinate conversion correctly and in keeping the structural and regulatory
questions separate — two things that are easy to get wrong and whose failures are
silent.

The most common misinterpretation is reading `isoform_status` as a statement
about conditions. If you take one thing from this vignette, take that: the
structural column describes which transcript carries a site, and the regulatory
column describes which condition detected it. They are both useful, and they are
not the same.

Suggestions and bug reports are welcome at
[github.com/Tarunchikatipalli6/m6Aswitch/issues](https://github.com/Tarunchikatipalli6/m6Aswitch/issues).

## Session Information {#session}

```{r session_info}
packageVersion("m6Aswitch")
sessionInfo()
```

[Back to Table of Contents](#table-of-contents)
